---
title: "QTA Lab 10: LLM-Assisted Coding"
format:
  gfm: default
editor: source
execute:
  echo: true
  message: false
  warning: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  results = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

## Learning Goals

In this lab, you will learn how to:

- define a qualitative coding task that can be applied at scale;
- write a transparent `quallmer` codebook;
- use `ellmer` syntax for model/provider settings;
- code leader speeches for credit claiming and blame attribution;
- compare coded claims across Labour and Conservative leaders;
- validate LLM-coded output against human judgements;
- optionally set up Ollama for local model use.

The substantive question is:

> Which Labour and Conservative leaders are more likely to claim credit, and which are more likely to attribute blame?

This is a useful task for LLM-assisted coding because the categories are interpretive. We are not only asking whether a speech contains words like "success" or "failure". We are asking whether the speaker uses the text to claim responsibility for a positive outcome, attribute responsibility for a negative outcome, both, or neither.

`quallmer` is by no means the only way to interact with LLMs from R. You can also use `ellmer` directly, provider-specific packages, web interfaces, or local tools. We use `quallmer` here because it encourages a structured qualitative coding workflow: define a codebook, request structured outputs, compare coding runs, validate against human judgements, and keep an audit trail.

## Start Here: Install And Test Ollama

The easiest way to try LLM-assisted coding without an API key is to run a small open-weight model locally with Ollama. This follows the setup used in the `quallmer` Ollama tutorial.

First, install the Ollama app from <https://ollama.com/download>. Start the app outside R. On Windows or macOS, you can usually do this by clicking the Ollama app icon. On macOS with Homebrew, you can also install and start it from the terminal:

```{bash install_ollama_macos, eval=FALSE}
brew install ollama
ollama serve
```

Then install the R package `rollama`, check that R can see Ollama, and download a small teaching model. The model `llama3.2:1b` is relatively small, so it is a good starting point for class. It is not necessarily the best model for research.

```{r setup_ollama_with_rollama, eval=FALSE}
install.packages("rollama")

library(rollama)

ping_ollama()

pull_model("llama3.2:1b")
```

Test the model from R:

```{r test_ollama_with_rollama, eval=FALSE}
rollama::query(
  "In one sentence, explain the difference between claiming credit and attributing blame.",
  model = "llama3.2:1b"
)
```

Check which models are installed using `rollama`:

```{r inspect_local_ollama_models_rollama}
if (requireNamespace("rollama", quietly = TRUE)) {
  rollama::list_models()
} else {
  message("Install rollama first with install.packages(\"rollama\").")
}
```

You can also check the same thing from the terminal:

```{bash list_ollama_models, eval=FALSE}
ollama list
```

Once a model is available, `quallmer` can use it through the model string `"ollama/model_name"`:

```{r quallmer_ollama_model_string, eval=FALSE}
model_name <- "ollama/llama3.2:1b"
```

Later in the lab, we will also show the lower-level `ellmer` syntax:

```{r ellmer_ollama_setup_preview, eval=FALSE}
local_chat <- ellmer::chat_ollama(
  model = "llama3.2:1b",
  params = ellmer::params(temperature = 0)
)
```

## Package Setup

`quallmer` organises the coding workflow: codebooks, coding runs, validation, comparison, and provenance. `ellmer` provides the model interface underneath. Live LLM calls require a working provider setup and usually an API key or a local Ollama model. In this lab, code that would contact an LLM is shown for you to inspect and run manually, but Quarto will skip it when rendering the document. This keeps the lab renderable even if you have not set up an API key or local model yet.

```{r check_llm_packages}
quallmer_installed <- requireNamespace("quallmer", quietly = TRUE)
ellmer_installed <- requireNamespace("ellmer", quietly = TRUE)

tibble::tibble(
  package = c("quallmer", "ellmer"),
  available = c(quallmer_installed, ellmer_installed)
)
```

If needed, install or update the packages before running live LLM calls.

```{r install_packages, eval=FALSE}
install.packages("ellmer")

# If quallmer is not yet available from your usual repository:
# install.packages("pak")
# pak::pak("SeraphineM/quallmer")
```

```{r load_packages}
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(ggplot2)
library(quallmer)
library(ellmer)

party_colors <- c(
  "Conservative" = "#003B73",
  "Labour" = "#B00020"
)
```

## Read The Leader Corpus

We use the same House of Commons leader corpus as Labs 6 and 7. The corpus contains up to 200 speeches by each Labour and Conservative party leader in each legislative period from the 47th Parliament, beginning with the 3 May 1979 general election, to the 58th Parliament, beginning with the 4 July 2024 general election.

Unlike Lab 6, we do not aggregate speeches into leader-period documents. Here the unit of analysis is one speech. That is important: credit claiming and blame attribution often occur in particular interventions, questions, or debate contributions.

```{r read_leader_corpus}
data_path <- "Data/hc_leader_period_sample_1979_2024.rds"

hoc_leaders <- readRDS(data_path) |>
  filter(party %in% c("Conservative", "Labour")) |>
  mutate(
    year = as.integer(format(date, "%Y")),
    speech_id = paste0("leader_", row_number()),
    party = factor(party, levels = c("Conservative", "Labour")),
    leader_period = paste(leader, election_period, sep = ": ")
  )

hoc_leaders |>
  count(party, leader, sort = TRUE)
```

For LLM-assisted coding, start small. We create a balanced teaching sample across leaders. In a research project, we would code more speeches and validate the output carefully.

```{r sample_leader_speeches_for_llm_coding}
llm_pool <- hoc_leaders |>
  filter(
    terms >= 80,
    !is.na(text),
    text != ""
  )

set.seed(20260710)

leader_llm <- llm_pool |>
  group_by(party, leader) |>
  mutate(random_order = runif(n())) |>
  arrange(random_order, .by_group = TRUE) |>
  slice_head(n = 4) |>
  ungroup() |>
  select(-random_order) |>
  mutate(
    text_for_coding = str_squish(text),
    text_for_coding = str_trunc(text_for_coding, width = 1800, side = "right")
  )

leader_llm |>
  select(speech_id, date, party, leader, election_period, agenda, text_for_coding) |>
  slice_head(n = 4)
```

The `text_for_coding` column is truncated because many models have context limits, and because short examples are easier to inspect. In a real project, truncation is a measurement decision. You would need to report it and test whether it changes the coding.

## The Coding Task

We will code whether each speech contains:

- a **credit claim**: the speaker claims that they, their party, their government, or a relevant actor delivered a positive outcome;
- a **blame attribution**: the speaker attributes a negative outcome, failure, harm, or problem to another actor;
- **both**: the speaker contrasts their own positive record with another actor's failure;
- **neither**: the speech does not contain a clear credit or blame claim.

This task is more interpretive than a dictionary count. A speech can mention "jobs", "waiting lists", or "inflation" without claiming credit or assigning blame. We are interested in the rhetorical move: who is made responsible for what?

## Codebook Categories

```{r define_credit_blame_labels}
claim_type_labels <- c(
  "credit_claim",
  "blame_attribution",
  "mixed_credit_and_blame",
  "neither"
)

target_actor_labels <- c(
  "own_party_or_government",
  "opposing_party_or_government",
  "public_body",
  "private_actor",
  "international_actor",
  "unclear_or_none"
)

policy_area_labels <- c(
  "economy",
  "public_services",
  "immigration",
  "foreign_affairs",
  "environment_energy",
  "constitutional_procedural",
  "other_or_unclear"
)
```

These labels are deliberately compact. If the codebook has too many categories, both human coders and LLMs will struggle to apply it consistently. If it has too few, we may lose the qualitative detail that made LLM-assisted coding useful.

## Structured Output Schema

The codebook tells the model how to think about the concept. The schema tells the model what the answer must look like. In the block below:

- `type_object()` says that each coded speech should be returned as one structured record;
- `type_boolean()` creates TRUE/FALSE fields for credit claiming and blame attribution;
- `type_enum()` restricts the model to predefined labels;
- `type_string()` allows short open-text answers where we want qualitative detail;
- `type_number()` asks for a numeric confidence score between 0 and 1.

This matters because we want output that can be inspected like qualitative coding, but also analysed like a data frame.

```{r define_credit_blame_schema}
credit_blame_schema <- ellmer::type_object(
  credit_claim = ellmer::type_boolean(
    "TRUE if the speaker claims credit for a positive outcome, improvement, achievement, or delivery."
  ),
  blame_attribution = ellmer::type_boolean(
    "TRUE if the speaker attributes a negative outcome, failure, harm, or problem to an actor."
  ),
  main_claim_type = ellmer::type_enum(
    claim_type_labels,
    "Overall classification of the speech."
  ),
  target_actor = ellmer::type_enum(
    target_actor_labels,
    "Actor mainly credited or blamed."
  ),
  target_name = ellmer::type_string(
    "Specific actor credited or blamed, if stated. Use 'unclear' if not stated."
  ),
  policy_area = ellmer::type_enum(
    policy_area_labels,
    "Main policy area of the credit or blame claim."
  ),
  confidence = ellmer::type_number(
    "Confidence in the coding decision from 0 to 1."
  ),
  evidence = ellmer::type_string(
    "Short quote or close paraphrase supporting the coding decision."
  )
)

credit_blame_schema
```

## Define The quallmer Codebook

The codebook is the measurement instrument. It should contain enough information for someone else to understand what the model was asked to do.

The `qlm_codebook()` object has four main parts:

- `name`: a short label for the coding task. This is useful when saving, comparing, or reporting coding runs.
- `role`: the perspective we ask the model to adopt. Here, we ask it to act like a careful qualitative coder of parliamentary speeches, not like a political commentator or general chatbot.
- `instructions`: the substantive coding rules. These define credit claiming, blame attribution, mixed cases, and cases that should be coded as neither. They also tell the model not to use outside knowledge about parties or leaders.
- `schema`: the structured output format we defined above. The instructions tell the model how to decide; the schema tells it how to return the decision.

Think of the codebook as the equivalent of a human-coder instruction sheet. If the codebook is vague, the model may still return confident answers, but those answers will be harder to interpret and validate.

```{r define_credit_blame_codebook}
credit_blame_codebook <- qlm_codebook(
  name = "HoC leader credit claiming and blame attribution",
  role = "You are a careful qualitative coder of UK parliamentary speeches.",
  instructions = paste(
    "Code the speech for credit claiming and blame attribution.",
    "Use only the information in the speech text.",
    "A credit claim says or implies that the speaker, their party, their government, or another actor delivered a positive outcome, improvement, achievement, or successful policy.",
    "A blame attribution says or implies that an actor caused, worsened, failed to address, or is responsible for a negative outcome, failure, harm, or public problem.",
    "Code mixed_credit_and_blame when the speech clearly contains both credit claiming and blame attribution.",
    "Code neither when the speech is descriptive, procedural, or does not clearly claim credit or attribute blame.",
    "Do not infer from party identity, leader identity, or outside knowledge.",
    "If several claims appear, code the most substantively developed claim.",
    "Evidence should be short and should justify both the main_claim_type and the target_actor."
  ),
  schema = credit_blame_schema
)

credit_blame_codebook
```

## A Concrete Example

The next block uses a fixed example from the leader corpus. This is better than drawing the example from the random teaching sample, because the human coding should clearly match the text students see.

```{r show_example_speech}
example_speech <- hoc_leaders |>
  filter(
    leader == "David Cameron",
    date == as.Date("2011-11-15"),
    str_detect(str_to_lower(text), fixed("we have delivered more apprenticeships"))
  ) |>
  slice_head(n = 1) |>
  mutate(text_for_coding = str_trunc(str_squish(text), width = 900))

example_speech |>
  select(date, party, leader, agenda, text_for_coding)
```

A plausible human coding might look like this:

```{r example_human_coding}
tibble::tibble(
  credit_claim = TRUE,
  blame_attribution = TRUE,
  main_claim_type = "mixed_credit_and_blame",
  target_actor = "opposing_party_or_government",
  target_name = "Labour government",
  policy_area = "economy",
  confidence = 0.85,
  evidence = "contrasts Labour's NEET record with the claim that 'we have delivered more apprenticeships'"
)
```

The important point is that the evidence field should point back to the text. Without evidence, it is difficult to tell whether the model made a defensible coding decision.

## A First Ollama Check With ellmer

Before using `quallmer`, it is useful to see the lower-level `ellmer` syntax. The code below creates a chat object for a local Ollama model and sends one message to it.

The chunk is not run automatically when rendering the lab. Run it manually once Ollama is installed, running, and the model has been downloaded.

```{r ellmer_ollama_example, eval=FALSE}
chat <- ellmer::chat_ollama(
  model = "llama3.2:1b",
  params = ellmer::params(temperature = 0)
)

chat$chat(
  "In one sentence, distinguish credit claiming from blame attribution."
)
```

## Code Speeches With Ollama And quallmer

Named character vectors are useful because `quallmer` stores names as `.id` values in the coded output.

```{r create_named_text_vector}
leader_texts <- leader_llm$text_for_coding
names(leader_texts) <- leader_llm$speech_id

head(names(leader_texts))
```

For classroom speed, we code a small balanced subset of speeches. The model used below is local, but it still takes time to run. That is why the live model call is shown but not run automatically during rendering.

```{r select_ollama_demo_texts}
ollama_demo_ids <- c(
  "leader_4898", "leader_5492", "leader_4975", "leader_5518",
  "leader_3939", "leader_3968", "leader_3927", "leader_3916"
)

ollama_demo_metadata <- leader_llm |>
  filter(speech_id %in% ollama_demo_ids) |>
  mutate(order = match(speech_id, ollama_demo_ids)) |>
  arrange(order) |>
  select(-order)

ollama_demo_texts <- ollama_demo_metadata$text_for_coding
names(ollama_demo_texts) <- ollama_demo_metadata$speech_id

ollama_demo_metadata |>
  select(speech_id, date, party, leader, agenda)
```

The next chunk is the actual `quallmer` call that sends the selected speeches to a local Ollama model for coding.

- `ollama_demo_texts` is the named vector of leader speeches.
- `credit_blame_codebook` contains the coding instructions and structured output schema.
- `model = "ollama/llama3.2"` tells `ellmer` to use the locally installed Ollama model `llama3.2`.
- `params = ellmer::params(temperature = 0)` makes the model output as stable as possible.
- `name` gives the coding run a clear label for provenance and later comparison.

```{r qlm_code_ollama_live, eval=FALSE}
coded_leaders_ollama <- qlm_code(
  ollama_demo_texts,
  credit_blame_codebook,
  model = "ollama/llama3.2",
  params = ellmer::params(temperature = 0),
  name = "leader_credit_blame_llama32_temperature0_demo"
)

coded_leaders_ollama
```

If you only have the smaller teaching model installed, change the model string to `"ollama/llama3.2:1b"`. It will run faster, but the output may be less reliable.

## Inspect Actual Ollama Output

The object below stores output from an actual local Ollama run using `llama3.2` with `temperature = 0`. This keeps the lab renderable while still letting us inspect real model output. If you rerun the previous chunk on your own machine, your output may differ slightly because local model versions and prompt handling can vary.

```{r load_actual_ollama_output}
ollama_coded_leaders <- tibble::tibble(
  .id = c(
    "leader_4898", "leader_5492", "leader_4975", "leader_5518",
    "leader_3939", "leader_3968", "leader_3927", "leader_3916"
  ),
  credit_claim = c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, FALSE),
  blame_attribution = c(TRUE, FALSE, TRUE, FALSE, TRUE, TRUE, FALSE, TRUE),
  main_claim_type = factor(
    c(
      "blame_attribution", "neither", "blame_attribution", "neither",
      "blame_attribution", "blame_attribution", "credit_claim",
      "blame_attribution"
    ),
    levels = claim_type_labels
  ),
  target_actor = factor(
    c(
      "unclear_or_none", "unclear_or_none", "opposing_party_or_government",
      "international_actor", "public_body", "private_actor", "public_body",
      "private_actor"
    ),
    levels = target_actor_labels
  ),
  target_name = c(
    "European friends and partners",
    "unspecified",
    "the Leader of the Opposition",
    "India authorities",
    "News International",
    "the Prime Minister's government or administration",
    "United States",
    "British arms"
  ),
  policy_area = factor(
    c(
      "foreign_affairs", "economy", "foreign_affairs", "foreign_affairs",
      "immigration", "public_services", "foreign_affairs", "foreign_affairs"
    ),
    levels = policy_area_labels
  ),
  confidence = c(0.8, 0.5, 0.8, 0.5, 0.8, 0.8, 0.8, 0.8),
  evidence = c(
    "has made it more difficult for us to get a deal",
    "The ability to roll out this vaccine",
    "That is the result of the decision that the Leader of the Opposition took last night.",
    "We will continue to work closely with the Indian authorities",
    "News International knew that phone hacking was widespread as long ago as 2007.",
    "He cannot even be in the same room as the doctors and nurses.",
    "We are grateful to President Obama for taking the decision and to the US special forces who carried it out.",
    "we are worried about the use of British arms for internal repression"
  )
)

ollama_coded_leaders |>
  count(main_claim_type, sort = TRUE)
```

## Join Codes Back To Metadata

The model output only contains the coding results and the `.id` for each speech. To ask substantive questions about parties and leaders, we need to join those codes back to the original metadata: date, party, leader, election period, agenda, and the speech text.

```{r join_codes_to_metadata}
coded_with_metadata <- ollama_coded_leaders |>
  left_join(
    leader_llm |>
      select(
        speech_id,
        date,
        year,
        party,
        leader,
        election_period,
        agenda,
        text_for_coding
      ),
    by = c(".id" = "speech_id")
  )

coded_with_metadata |>
  select(
    .id,
    party,
    leader,
    election_period,
    main_claim_type,
    target_actor,
    policy_area,
    confidence
  ) |>
  slice_head(n = 10)
```

## Compare Parties And Leaders

The next table asks the substantive question directly: in this coded sample, which leaders have higher estimated rates of credit claiming and blame attribution?

```{r summarise_claims_by_leader}
leader_claim_rates <- coded_with_metadata |>
  group_by(party, leader) |>
  summarise(
    n = n(),
    credit_rate = mean(credit_claim),
    blame_rate = mean(blame_attribution),
    mixed_rate = mean(main_claim_type == "mixed_credit_and_blame"),
    .groups = "drop"
  ) |>
  arrange(party, desc(blame_rate), desc(credit_rate))

leader_claim_rates
```

```{r plot_credit_blame_by_leader}
claim_plot_data <- leader_claim_rates |>
  pivot_longer(
    cols = c(credit_rate, blame_rate),
    names_to = "claim_measure",
    values_to = "rate"
  ) |>
  mutate(
    claim_measure = factor(
      recode(
        claim_measure,
        credit_rate = "Credit claiming",
        blame_rate = "Blame attribution"
      ),
      levels = c("Credit claiming", "Blame attribution")
    ),
    leader_label = paste(leader, party, sep = "\n")
  )

leader_levels <- leader_claim_rates |>
  arrange(party, leader) |>
  mutate(leader_label = paste(leader, party, sep = "\n")) |>
  pull(leader_label)

claim_plot_data |>
  mutate(
    leader_label = factor(leader_label, levels = rev(leader_levels))
  ) |>
  ggplot(aes(x = claim_measure, y = leader_label)) +
  geom_tile(
    aes(fill = party),
    alpha = 0.28,
    color = "white",
    linewidth = 1.2,
    width = 0.95,
    height = 0.85
  ) +
  geom_text(
    aes(label = scales::percent(rate, accuracy = 1)),
    size = 5,
    fontface = "bold",
    color = "grey15"
  ) +
  scale_fill_manual(values = party_colors) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Party"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(lineheight = 0.95),
    legend.position = "bottom"
  )
```

This plot is based on a small Ollama-coded sample, so we should not interpret the substantive pattern as evidence. The workflow is the key point: code speeches, join codes to metadata, then compare leaders.

## Inspect The Evidence

The evidence field matters because it makes the coding easier to audit. If the evidence does not support the label, the model may have guessed, over-generalised, or relied on irrelevant cues.

```{r inspect_coded_evidence}
coded_with_metadata |>
  filter(main_claim_type != "neither") |>
  select(
    date,
    party,
    leader,
    agenda,
    main_claim_type,
    target_actor,
    policy_area,
    evidence
  ) |>
  slice_head(n = 8)
```

## Check Output Consistency

Structured output makes it easier to check whether the model contradicted itself. Here, `main_claim_type` should agree with the two Boolean fields:

- `credit_claim` means credit is TRUE and blame is FALSE;
- `blame_attribution` means blame is TRUE and credit is FALSE;
- `mixed_credit_and_blame` means both are TRUE;
- `neither` means both are FALSE.

```{r check_ollama_output_consistency}
checked_ollama_output <- ollama_coded_leaders |>
  mutate(
    expected_main_claim_type = case_when(
      credit_claim & blame_attribution ~ "mixed_credit_and_blame",
      credit_claim ~ "credit_claim",
      blame_attribution ~ "blame_attribution",
      TRUE ~ "neither"
    ),
    internally_consistent = main_claim_type == expected_main_claim_type
  )

checked_ollama_output |>
  select(
    .id,
    credit_claim,
    blame_attribution,
    main_claim_type,
    expected_main_claim_type,
    internally_consistent
  )
```

If this check fails, do not simply fix the data silently. Inspect the speech, revise the codebook if needed, and rerun the model. In a real project, internal consistency is only the first check. We still need human validation.

## Validate Against Human Coding

No LLM coding project is complete without validation. A practical strategy is to human-code a stratified sample across parties, leaders, and model-predicted labels. This matters here because credit and blame may look different in government and opposition, and across leaders with different rhetorical styles.

Here we create a small mock human-coded set. In your project, this should come from actual human coding.

```{r create_mock_human_gold}
human_gold <- ollama_coded_leaders |>
  slice_head(n = 12) |>
  select(
    .id,
    credit_claim,
    blame_attribution,
    main_claim_type,
    target_actor,
    policy_area
  ) |>
  mutate(
    main_claim_type = case_when(
      row_number() %in% c(3, 8) ~ "neither",
      TRUE ~ main_claim_type
    ),
    credit_claim = main_claim_type %in% c("credit_claim", "mixed_credit_and_blame"),
    blame_attribution = main_claim_type %in% c("blame_attribution", "mixed_credit_and_blame")
  )

human_gold
```

With real `quallmer` objects, convert the human data and validate.

```{r qlm_validate_live, eval=FALSE}
human_coded <- qlm_humancoded(
  human_gold,
  name = "expert_human_coder",
  codebook = credit_blame_codebook,
  texts = leader_texts,
  metadata = list(
    coder_training = "Instructor-coded validation sample",
    date = Sys.Date()
  )
)

qlm_validate(
  coded_leaders_ollama,
  human_coded,
  by = main_claim_type,
  level = "nominal",
  average = "none"
)
```

The runnable example below computes a simple validation table from the stored Ollama output.

```{r validate_example_output}
example_validation <- ollama_coded_leaders |>
  inner_join(
    human_gold |>
      rename(
        human_credit_claim = credit_claim,
        human_blame_attribution = blame_attribution,
        human_main_claim_type = main_claim_type,
        human_target_actor = target_actor,
        human_policy_area = policy_area
      ),
    by = ".id"
  ) |>
  mutate(
    claim_type_correct = main_claim_type == human_main_claim_type,
    credit_correct = credit_claim == human_credit_claim,
    blame_correct = blame_attribution == human_blame_attribution
  )

example_validation |>
  count(human_main_claim_type, main_claim_type)

example_validation |>
  summarise(
    claim_type_accuracy = mean(claim_type_correct),
    credit_accuracy = mean(credit_correct),
    blame_accuracy = mean(blame_correct),
    n = n()
  )
```

## Provenance And Reporting

`quallmer` keeps metadata about coding runs. This is useful for reproducibility: model, codebook, parameters, batch settings, and parent runs can all be tracked.

```{r qlm_trail_live, eval=FALSE}
trail <- qlm_trail(coded_leaders_ollama)

trail

# In the current version of quallmer, the export destination is called `file`.
qlm_trail_export(
  trail,
  file = "Labs/Lab_10/leader_credit_blame_coding_trail.json"
)
```

For your own work, save the codebook, model settings, prompts, validation data, and final coded output. The exact text of the codebook matters as much as the model name.

```{r create_report_skeleton}
report_checklist <- tibble::tribble(
  ~item, ~why_it_matters,
  "Concept and codebook", "Defines credit claiming and blame attribution",
  "Structured output schema", "Defines allowed labels and scales",
  "Leader corpus and sampling", "Defines the population of speeches being compared",
  "Model/provider/version", "Affects reproducibility and performance",
  "Local model or API model", "Affects reproducibility, privacy, cost, and access",
  "Temperature and other parameters", "Affects stability and variation",
  "Input truncation or chunking rules", "Affects what the model can see",
  "Human validation sample design", "Shows how accuracy was assessed",
  "Validation metrics by label and leader", "Reveals where the model works or fails",
  "Examples of errors", "Reveals conceptual and practical failure modes"
)

report_checklist
```

## Extensions

Useful extensions for a final project:

- compare local Ollama coding to an API model;
- test whether results change when leader and party metadata are included or excluded;
- compare credit/blame rates between government and opposition leaders;
- code the same speeches with two versions of the codebook;
- compare LLM-coded credit/blame claims with a simpler dictionary baseline;
- use human-coded cases to refine codebook instructions before scaling up.

## Optional: Using OpenAI Instead Of Ollama

Ollama is useful because it can run locally without an API key. For larger or more reliable models, you may want to use an API provider instead. With OpenAI, create an API key in the OpenAI platform dashboard and store it as an environment variable called `OPENAI_API_KEY`. OpenAI's quickstart explains this setup, and the API-key help page explains where keys are created and why they should not be shared.

```{r openai_api_key_setup, eval=FALSE}
# In your terminal, not inside the Quarto document:
# export OPENAI_API_KEY="your_api_key_here"
```

Once the key is available to R, the `ellmer` syntax is very similar:

```{r ellmer_openai_example, eval=FALSE}
chat <- ellmer::chat_openai(
  model = "gpt-4o-mini",
  params = ellmer::params(temperature = 0)
)

chat$chat(
  "In one sentence, distinguish credit claiming from blame attribution."
)
```

The corresponding `quallmer` call only changes the `model` argument:

```{r qlm_code_openai_live, eval=FALSE}
coded_leaders_openai <- qlm_code(
  ollama_demo_texts,
  credit_blame_codebook,
  model = "openai/gpt-4o-mini",
  params = ellmer::params(temperature = 0),
  name = "leader_credit_blame_gpt4o_mini_temperature0_demo"
)

coded_leaders_openai
```

Official OpenAI guidance: <https://platform.openai.com/docs/quickstart/make-your-first-api-request> and <https://help.openai.com/en/articles/4936850-where-do-i-find-my-openai-api-key>.

## Practice Exercises

1. Draft concise coding instructions for a binary task: whether a speech contains blame attribution. Store them in an object called `binary_blame_instructions`.

```{r exercise_binary_codebook}
# Your answer here
```

2. Define a structured schema for the binary blame task with fields `blame_attribution`, `confidence`, and `evidence`. Then combine the instructions and schema in a `quallmer` codebook called `binary_blame_codebook`.

```{r exercise_binary_schema}
# Your answer here
```

3. Show the `qlm_code()` call you would use to code the first five leader speeches with this binary blame codebook.

```{r exercise_binary_qlm_code, eval=FALSE}
# Your answer here
```

4. Create a small mock human-coded data frame with `.id` and `blame_attribution` for the first eight speeches.

```{r exercise_mock_human_codes}
# Your answer here
```

5. Show how you would convert those human codes with `qlm_humancoded()` and validate an LLM-coded result with `qlm_validate()`.

```{r exercise_validate_binary_codes, eval=FALSE}
# Your answer here
```

6. Using `ollama_coded_leaders`, calculate the share of speeches with credit claims and blame attributions by `leader`.

```{r exercise_claim_share_by_leader}
# Your answer here
```

7. Inspect three speeches where the Ollama output labels the claim type as `neither`. What information would you need to decide whether that label is correct?

```{r exercise_inspect_neither}
# Your answer here
```

8. Reflection: what should be included in a methods section when reporting LLM-assisted coding of credit claiming and blame attribution with `quallmer` and `ellmer`?

```{r exercise_methods_reflection}
# Your notes here
```
