---
title: "QTA Lab 09 Answers: Zero-Shot Classification with NLI Models"
format:
  gfm: default
editor: source
execute:
  echo: true
  message: false
  warning: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  results = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

## Learning Goals

In this lab, you will learn how to:

- explain why transformer models make transfer learning useful for text classification;
- define a zero-shot classification task using labels and natural-language hypotheses;
- use Natural Language Inference (NLI) logic for document classification;
- compare model-style labels against a weak keyword baseline;
- create a small validation sample for inspecting model-generated labels;
- understand how the optional transformer extension differs from the Part 1 teaching workflow.

## Conceptual Overview

In traditional supervised machine learning, we usually need labelled training examples for the exact task we want to solve. Classical text classifiers often start from a document-feature matrix, so they must learn both the relevant language patterns and the classification task from the labelled examples we provide.

Transformer models change this starting point. They are pre-trained on large text corpora and learn contextual representations of language: the same word can be represented differently depending on the surrounding text. This is why they are useful for transfer learning. Instead of learning everything from our House of Commons sample, we can reuse a model that already contains general language representations.

BERT-NLI is one specific kind of transformer-based model. It is not the only BERT-style model, and BERT is not a single method for all tasks. Other models include BERT, RoBERTa, DeBERTa, Longformer, domain-specific models, and multilingual models. BERT-NLI is especially useful for this lab because it has also been trained on Natural Language Inference (NLI), which gives it task knowledge about comparing two pieces of text.

NLI models are trained to compare two input texts: a `context` and a `hypothesis`. The task is to determine whether the hypothesis is `true`, `false`, or `neutral` given the context. In other terminology, these classes are often called `entailment`, `contradiction`, and `neutral`.

For document classification, we turn a label into a hypothesis:

- context: a House of Commons speech;
- candidate label: `health`;
- hypothesis: "This speech is about health policy."

If the NLI model assigns a high `true` or entailment probability to that hypothesis, we treat the speech as likely to belong to the `health` category. This is not about whether the hypothesis is objectively true in the world. It is about whether the speech text supports that hypothesis.

This is powerful because we can classify documents without collecting hundreds or thousands of labelled examples first. It is also risky because the model choice, labels, hypotheses, corpus fit, and validation design matter enormously.

## Load Packages

```{r load_packages}
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(ggplot2)
library(quanteda)
library(caret)
library(reticulate)

party_colors <- c(
  "Conservative" = "#003B73",
  "Labour" = "#B00020"
)
```

## Part 1: Classification Workflow Without Transformers

## Read And Prepare The HoC Data

```{r read_hoc_sample}
data_path <- "Data/hc_sample_1945_2025.rds"

hoc <- readRDS(data_path)

hoc <- hoc |>
  mutate(
    year = as.integer(format(date, "%Y")),
    speech_id = paste0("hoc_", row_number()),
    party_group = case_when(
      party %in% c("Conservative", "Labour") ~ party,
      TRUE ~ "Other"
    ),
    party_group = factor(party_group, levels = c("Conservative", "Labour", "Other"))
  )
```

We use a small sample of recent Conservative and Labour speeches. Keeping
the party comparison simple makes the examples easier to inspect. Transformer
inference is much slower than ordinary dictionary or DFM methods, so start
small and scale up only after the workflow is correct.

```{r sample_hoc_for_nli}
nli_pool <- hoc |>
  filter(
    year >= 1997,
    party_group %in% c("Conservative", "Labour"),
    terms >= 80,
    !is.na(text),
    text != ""
  )

set.seed(20260710)

hoc_nli <- nli_pool |>
  slice_sample(n = min(80, nrow(nli_pool)))

hoc_nli |>
  count(party_group, sort = TRUE)
```

Most transformer models have maximum context lengths. We therefore create a shortened text field for the lab of up to 220 tokens. In a real project, you would decide whether to truncate, split long documents into chunks, classify each chunk, or use a long-context model.

The function below keeps the first `max_words` words of each speech. This
is not a substantive claim about which part of the speech matters most. It
is a practical way to keep the lab fast and to mimic the context limits
that transformer models often impose.

```{r truncate_speeches}
truncate_words <- function(x, max_words = 220) {
  words <- str_split(str_squish(x), "\\s+")[[1]]
  paste(head(words, max_words), collapse = " ")
}

hoc_nli <- hoc_nli |>
  mutate(text_for_model = map_chr(text, truncate_words))

hoc_nli |>
  select(speech_id, date, speaker, party, agenda, text_for_model) |>
  slice_head(n = 3)
```

## Define Candidate Labels

Good zero-shot labels should be clear, mutually intelligible, and close to the concepts you actually want to measure. We define labels and then write label-specific hypotheses.

```{r define_candidate_labels}
label_descriptions <- tibble(
  label = c(
    "economy",
    "health",
    "security",
    "environment",
    "education",
    "housing",
    "procedural"
  ),
  hypothesis = c(
    "This speech is about the economy, taxation, public spending, jobs, or business.",
    "This speech is about health policy, the NHS, hospitals, patients, or social care.",
    "This speech is about security, defence, policing, crime, borders, or terrorism.",
    "This speech is about climate change, energy, carbon emissions, or environmental policy.",
    "This speech is about schools, universities, teachers, students, or education policy.",
    "This speech is about housing, rent, homes, tenants, planning, or homelessness.",
    "This speech is mainly procedural parliamentary business rather than substantive policy."
  )
)

label_descriptions
```

This table does two things. The `label` column gives the short category
names that will appear in our output. The `hypothesis` column states what
each label means in ordinary language. In a real NLI model, these
hypotheses are compared to each speech. In Part 1, we keep the
hypotheses visible because they are still the codebook for the task.

## A Weak Keyword Baseline

Before using a transformer, build a simple baseline. This gives you something to compare against and helps reveal whether the zero-shot model is doing something useful.

These are not gold-standard labels. They are weak labels.

The code below has three steps. First, `keyword_patterns` defines one
regular expression per label. Second, `map_dfc()` applies each pattern to
every speech and stores the counts in separate columns. Third, the
`rowwise()` block selects the label with the highest keyword count for
each speech.

```{r create_keyword_baseline}
keyword_patterns <- c(
  economy = "\\b(economy|economic|tax|taxes|budget|inflation|growth|business|businesses|jobs|wages|spending)\\b",
  health = "\\b(health|nhs|hospital|hospitals|patient|patients|care|doctor|doctors|nurse|nurses)\\b",
  security = "\\b(security|defence|police|crime|criminal|terrorism|terrorist|border|borders|army|military)\\b",
  environment = "\\b(climate|environment|environmental|carbon|emissions|energy|green|renewable|pollution)\\b",
  education = "\\b(education|school|schools|teacher|teachers|student|students|university|universities|pupils)\\b",
  housing = "\\b(housing|housebuilding|homes|rent|rents|tenant|tenants|landlord|homeless|homelessness|planning)\\b",
  procedural = "\\b(order|motion|committee|clause|amendment|amendments|bill|division|standing order)\\b"
)

keyword_scores <- map_dfc(
  keyword_patterns,
  ~ str_count(str_to_lower(hoc_nli$text_for_model), regex(.x))
)

keyword_baseline <- bind_cols(
  hoc_nli |>
    select(speech_id, date, year, speaker, party, party_group, agenda),
  keyword_scores
) |>
  rowwise() |>
  mutate(
    max_keyword_score = max(c_across(all_of(names(keyword_patterns)))),
    weak_label = names(keyword_patterns)[
      which.max(c_across(all_of(names(keyword_patterns))))
    ],
    weak_label = if_else(max_keyword_score > 0, weak_label, "unlabelled")
  ) |>
  ungroup()

keyword_baseline |>
  count(weak_label, sort = TRUE)
```

The output tells us which labels the keyword rules assign most often.
This is only a rough baseline. Its main purpose is to give us a transparent
comparison point for later model-style labels.

## Keep The First Workflow Simple

For the first part of the lab, we deliberately do **not** run the
transformer model. This keeps the workflow fast, reproducible, and easier
to understand.

The objects we create below have the same shape as zero-shot classifier
output: one row per speech-label combination and one score for each
candidate label. The scores come from the keyword baseline, not from an
NLI model. This lets us learn the workflow before complicating things by relying on Python, model downloads, etc.

```{r set_part_one_mode}
run_nli <- FALSE
transformer_available <- FALSE

run_nli
```

For your own use, start with the simple workflow. Once the data
structure, labels, hypotheses, validation checks, and reporting logic are
clear, run the optional transformer extension in Part 2.

```{r define_nli_model}
model_name <- "MoritzLaurer/deberta-v3-base-zeroshot-v1.1-all-33"

hypothesis_template <- "This speech is about {}."
```

## A Teaching Classifier

The function below is a teaching device. It takes one speech and a set of
candidate labels. For each label, it counts how often the matching keyword
pattern appears in the speech. It then rescales those counts so that the
label scores add up to one.

This is **not** a zero-shot NLI model. It is useful because it returns the
same kind of table that a zero-shot model would return: labels ranked by
score. This means all later steps--choosing the top label, inspecting
disagreements, checking sensitivity, and preparing validation samples--can
be learned without running a transformer.

The `classify_zero_shot()` function does four things:

1. It loops over the candidate labels, such as `economy`, `health`, or
   `housing`.
2. For each label, it counts how often the matching keyword pattern
   appears in the speech.
3. If no keywords are found for any label, it assigns equal scores to all
   labels so that the function still returns a usable result.
4. It rescales the scores so they sum to one and sorts the labels from
   highest to lowest score.

The `hypothesis_template` argument is included so that the function has
the same structure as the real NLI classifier in Part 2. In Part 1, the
keyword-based teaching classifier does not actually use the template.

```{r define_zero_shot_classifier}
classify_zero_shot <- function(text, labels, hypothesis_template) {
  raw_scores <- map_dbl(labels, function(label) {
    if (label %in% names(keyword_patterns)) {
      str_count(str_to_lower(text), regex(keyword_patterns[[label]]))
    } else {
      0
    }
  })

  if (sum(raw_scores) == 0) {
    raw_scores <- rep(1, length(labels))
  }

  tibble(
    label = labels,
    score = raw_scores / sum(raw_scores)
  ) |>
    arrange(desc(score))
}
```

## Classify A Small Batch Of Speeches

We begin with 25 speeches. In Part 1 this is fast because we use keyword
scores. With a real transformer model, the same code can become much
slower, so it is good practice to start with a small batch.

The code below loops over speech IDs and speech texts at the same time.
For each speech, it returns a table of label scores. `map2_dfr()` then
binds all those small tables into one larger table.

```{r run_zero_shot_classification}
nli_batch <- hoc_nli |>
  slice_head(n = 25)

label_scores <- map2_dfr(
  nli_batch$speech_id,
  nli_batch$text_for_model,
  function(id, speech_text) {
    classify_zero_shot(
      text = speech_text,
      labels = label_descriptions$label,
      hypothesis_template = hypothesis_template
    ) |>
      mutate(speech_id = id, .before = 1)
  }
)

label_scores |>
  arrange(speech_id, desc(score)) |>
  group_by(speech_id) |>
  slice_head(n = 3) |>
  ungroup()
```

Each speech now has several candidate labels. For most analyses we need a
single predicted label per speech. The next block keeps the highest-scoring
label for each speech and joins the speech metadata back in.

```{r create_predictions}
predictions <- label_scores |>
  group_by(speech_id) |>
  slice_max(order_by = score, n = 1, with_ties = FALSE) |>
  ungroup() |>
  rename(
    predicted_label = label,
    predicted_score = score
  ) |>
  left_join(
    nli_batch |>
      select(speech_id, date, year, speaker, party, party_group, agenda, text_for_model),
    by = "speech_id"
  )

predictions |>
  select(speech_id, party_group, agenda, predicted_label, predicted_score) |>
  arrange(desc(predicted_score))
```

## Compare With Weak Labels

The keyword labels are not ground truth, but disagreements are useful. They tell us which speeches need closer inspection and where the model or the dictionary may be interpreting the task differently.

In Part 1 the "classifier" and the weak baseline are closely related, so
this comparison is mainly about learning the validation workflow. In Part
2, the same comparison becomes more interesting because the NLI model may
label a speech differently from the keyword baseline.

```{r compare_nli_to_keyword_labels}
comparison <- predictions |>
  left_join(
    keyword_baseline |>
      select(speech_id, weak_label, max_keyword_score),
    by = "speech_id"
  ) |>
  mutate(agrees_with_weak_label = predicted_label == weak_label)

comparison |>
  count(predicted_label, weak_label, sort = TRUE)
```

```{r inspect_disagreements}
comparison |>
  filter(!agrees_with_weak_label) |>
  select(
    speech_id,
    party_group,
    agenda,
    predicted_label,
    predicted_score,
    weak_label,
    max_keyword_score,
    text_for_model
  ) |>
  slice_head(n = 5)
```

When inspecting disagreements, read the speech text and ask: is the
predicted label substantively plausible, did the keyword baseline fire for
an incidental reason, or is the case genuinely mixed?

## Visualise Predicted Labels

```{r plot_zero_shot_labels_by_party}
comparison |>
  count(party_group, predicted_label) |>
  ggplot(
    aes(
      x = predicted_label,
      y = n,
      fill = party_group
    )
  ) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = party_colors) +
  labs(
    x = NULL,
    y = "Number of speeches",
    fill = "Party group"
  ) +
  theme_minimal()
```

## Create A Validation Sample

The most important habit in this lab is not to trust the predicted labels
automatically. We need a small set of cases that a human can inspect.

The code below keeps two high-scoring cases per predicted label. This is a
simple validation sample, not a full validation design. In a research
project, we would also include low-confidence cases, disagreements, and
important subgroups such as party or period.

```{r create_validation_sample}
validation_sample <- comparison |>
  group_by(predicted_label) |>
  slice_max(order_by = predicted_score, n = 2, with_ties = FALSE) |>
  ungroup() |>
  select(
    speech_id,
    predicted_label,
    predicted_score,
    weak_label,
    party_group,
    agenda,
    text_for_model
  )

validation_sample
```

## Part 2: Running A Real Zero-Shot NLI Model

The first part of the lab taught the workflow without requiring Python or
model downloads. The optional extension below runs the same kind of task
with a BERT/DeBERTa-style NLI model through Python's Hugging Face
`transformers` library. This is one transformer model choice, not the only
way to use transformers for classification.

This is more advanced for three reasons. First, R has to talk to Python
through `reticulate`. Second, the required Python packages must be
installed in the same Python environment that R is using. Third, the model
may need to be downloaded the first time you run it.

### Install Python Dependencies

We use a project-level virtual environment called `.venv`. The command
below installs the transformer packages into that environment. Run it in
the terminal, not inside R.

```{bash install_transformer_dependencies, eval=FALSE}
uv venv --python 3.13 .venv
uv pip install --python .venv/bin/python transformers torch sentencepiece accelerate
```

If you run the command from inside `Labs/Lab_9`, use the path to the
course-level environment:

```{bash install_transformer_dependencies_from_lab_folder, eval=FALSE}
uv pip install --python ../../.venv/bin/python transformers torch sentencepiece accelerate
```

### Connect R To The Python Environment

The next block tells `reticulate` which Python to use. This must happen
before Python has been initialized in the R session. If `reticulate` has
already chosen a different Python, restart R and run the lab from the top.

```{r setup_reticulate_for_nli, eval=FALSE}
library(reticulate)

python_candidates <- c(
  ".venv/bin/python",
  "../../.venv/bin/python"
)

python_path <- python_candidates[file.exists(python_candidates)][1]

if (!is.na(python_path)) {
  use_python(python_path, required = TRUE)
}

py_config()
```

### Check Whether The Transformer Stack Is Available

This check only asks whether the Python packages are available. It does
not yet download or run the model. The actual model is specified below by
`model_name`.

```{r check_transformer_environment, eval=FALSE}
run_nli <- TRUE

transformer_available <- py_module_available("transformers") &&
  py_module_available("torch")

transformer_available
```

### Define The Real NLI Classifier

This function sends a speech to the transformer pipeline as the
`context`. For each candidate label, the model evaluates whether the
context makes the label hypothesis `true` or entailed. The output is
deliberately made to look like the Part 1 output: a table with `label`
and `score`.

We define `model_name` and `hypothesis_template` again here so that this
advanced section can be run on its own after the setup chunks.

```{r define_real_nli_classifier, eval=FALSE}
model_name <- "MoritzLaurer/deberta-v3-base-zeroshot-v1.1-all-33"
hypothesis_template <- "This speech is about {}."

transformers <- import("transformers")

zero_shot_pipeline <- transformers$pipeline(
  task = "zero-shot-classification",
  model = model_name,
  device = -1L
)

classify_with_nli <- function(text, labels, hypothesis_template) {
  result <- zero_shot_pipeline(
    text,
    candidate_labels = as.list(labels),
    hypothesis_template = hypothesis_template,
    multi_label = TRUE
  )

  tibble(
    label = unlist(result$labels),
    score = as.numeric(result$scores)
  )
}
```

### Try The Real Model On A Very Small Batch

Start with a few speeches. The first run may take longer because the model
has to be downloaded and loaded into memory.

```{r run_real_nli_small_batch, eval=FALSE}
real_nli_batch <- hoc_nli |>
  slice_head(n = 5)

real_label_scores <- map2_dfr(
  real_nli_batch$speech_id,
  real_nli_batch$text_for_model,
  function(id, speech_text) {
    classify_with_nli(
      text = speech_text,
      labels = label_descriptions$label,
      hypothesis_template = hypothesis_template
    ) |>
      mutate(speech_id = id, .before = 1)
  }
)

real_label_scores |>
  arrange(speech_id, desc(score)) |>
  group_by(speech_id) |>
  slice_head(n = 3) |>
  ungroup()
```

## Reporting Checklist

When reporting zero-shot or weakly supervised classification, document:

- model name and version;
- labels and full hypothesis templates;
- whether classification was single-label or multi-label;
- truncation or chunking decisions;
- validation sample design;
- performance by label and by relevant subgroups;
- examples of false positives and false negatives;
- any use of weak labels or pseudo-labels.

## Practice Exercises

1. Add a new candidate label called `migration` with an appropriate hypothesis.

```{r exercise_add_migration_label}
label_descriptions_expanded <- bind_rows(
  label_descriptions,
  tibble(
    label = "migration",
    hypothesis = "This speech is about immigration, asylum, refugees, migrants, or border control."
  )
)

label_descriptions_expanded
```

2. Add a keyword pattern for `migration` to the weak baseline and inspect how many sampled speeches it labels.

```{r exercise_migration_keyword_baseline}
keyword_patterns_expanded <- c(
  keyword_patterns,
  migration = "\\b(immigration|immigrant|immigrants|migration|migrant|migrants|asylum|refugee|refugees|visa|visas)\\b"
)

keyword_scores_expanded <- map_dfc(
  keyword_patterns_expanded,
  ~ str_count(str_to_lower(hoc_nli$text_for_model), regex(.x))
)

keyword_baseline_expanded <- bind_cols(
  hoc_nli |>
    select(speech_id, date, year, speaker, party, party_group, agenda),
  keyword_scores_expanded
) |>
  rowwise() |>
  mutate(
    max_keyword_score = max(c_across(all_of(names(keyword_patterns_expanded)))),
    weak_label = names(keyword_patterns_expanded)[
      which.max(c_across(all_of(names(keyword_patterns_expanded))))
    ],
    weak_label = if_else(max_keyword_score > 0, weak_label, "unlabelled")
  ) |>
  ungroup()

keyword_baseline_expanded |>
  count(weak_label, sort = TRUE)
```

3. Run the Part 1 classification workflow for 10 speeches using the expanded label set.

```{r exercise_classify_with_migration}
migration_batch <- hoc_nli |>
  slice_head(n = 10)

migration_scores <- map2_dfr(
  migration_batch$speech_id,
  migration_batch$text_for_model,
  function(id, speech_text) {
    classify_zero_shot(
      text = speech_text,
      labels = label_descriptions_expanded$label,
      hypothesis_template = hypothesis_template
    ) |>
      mutate(speech_id = id, .before = 1)
  }
)

migration_predictions <- migration_scores |>
  group_by(speech_id) |>
  slice_max(order_by = score, n = 1, with_ties = FALSE) |>
  ungroup() |>
  rename(
    predicted_label = label,
    predicted_score = score
  ) |>
  left_join(
    migration_batch |>
      select(speech_id, party_group, agenda, text_for_model),
    by = "speech_id"
  )

migration_predictions |>
  select(speech_id, party_group, agenda, predicted_label, predicted_score)
```

4. Create a validation sample with at least one speech per predicted label.

```{r exercise_validation_sample}
migration_predictions |>
  group_by(predicted_label) |>
  slice_max(order_by = predicted_score, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(
    speech_id,
    predicted_label,
    predicted_score,
    party_group,
    agenda,
    text_for_model
  )
```

5. Identify three speeches where the predicted label and weak keyword label disagree. What should a human coder inspect?

```{r exercise_inspect_disagreements}
expanded_comparison <- migration_predictions |>
  left_join(
    keyword_baseline_expanded |>
      select(speech_id, weak_label, max_keyword_score),
    by = "speech_id"
  ) |>
  mutate(agrees_with_weak_label = predicted_label == weak_label)

expanded_comparison |>
  filter(!agrees_with_weak_label) |>
  select(
    speech_id,
    predicted_label,
    predicted_score,
    weak_label,
    max_keyword_score,
    agenda,
    text_for_model
  ) |>
  slice_head(n = 3)

# A human coder should inspect whether the speech is genuinely about the
# predicted label, whether the weak keyword match is incidental, whether the
# speech contains multiple topics, and whether truncation removed relevant text.
```

6. Write a short codebook entry for one label, including positive examples, negative examples, and borderline cases.

```{r exercise_codebook_entry}
# Example codebook entry for "migration":
#
# Code 1 if the speech substantively discusses immigration, asylum, refugees,
# migrants, visas, border control, or migration policy.
#
# Positive examples:
# - a speech about asylum processing or refugee resettlement;
# - a speech about immigration rules, visas, or border enforcement;
# - a speech about the social or economic effects of migration.
#
# Negative examples:
# - a speech that only mentions a migrant background as biographical detail;
# - a speech about national security where borders are mentioned only in passing;
# - a procedural amendment with no substantive migration content.
#
# Borderline cases:
# - speeches about trafficking may be migration, security, or both;
# - speeches about labour shortages may be economy unless migration policy is a
#   central part of the argument.
```

7. Reflection: how does NLI-based zero-shot classification differ from the Naive Bayes classifier in Lab 5? When would you still prefer human-labelled training data?

```{r exercise_transfer_learning_reflection}
# In Lab 5, the Naive Bayes classifier learned from labelled examples in the
# target dataset. It needed training labels for the specific task, and its
# features came from the local DFM.
#
# NLI-based zero-shot classification uses a model that has already learned a
# transferable inference task. Instead of training on local examples, we express
# each label as a hypothesis and ask whether the document entails it.
#
# I would still prefer human-labelled training data when the categories are
# subtle, high-stakes, domain-specific, or likely to be confused by rhetoric,
# irony, parliamentary procedure, or multi-topic speeches. Human labels are also
# essential for validation, even when the first round of labels comes from an NLI
# model.
```
