---
title: "QTA Lab 04: Dictionary methods with House of Commons speeches"
format:
  gfm: default
editor: source
execute:
  echo: true
  message: false
  warning: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  results = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

## Learning goals

In this lab, you will learn how to:

- apply off-the-shelf dictionaries to House of Commons speeches;
- calculate dictionary-based rates rather than raw counts;
- compare two sentiment dictionaries;
- build, apply, and adapt a small economic insecurity dictionary;
- inspect false positives and false negatives using KWIC;
- use static embeddings and LLM prompts as sources of candidate terms;
- report dictionary choices transparently.

Dictionary methods are transparent: we know which words are counted. That is their great strength. Their weakness is that word lists rarely map perfectly onto a concept, especially in political language where the same word can mean different things in different contexts. Today we use economic insecurity as the running example.

## Load packages

If you have not installed `quanteda.sentiment`, you may need to install it from GitHub:

```{r install_packages}
# install.packages("remotes")
# remotes::install_github("quanteda/quanteda.sentiment")
```

```{r load_packages}
library(dplyr)
library(stringr)
library(ggplot2)
library(quanteda)
library(quanteda.sentiment)
```

## Read the House of Commons sample

```{r read_hoc_sample}
data_path <- "Data/hc_sample_1945_2025.rds"

hoc <- readRDS(data_path)

hoc <- hoc |>
  mutate(year = as.integer(format(date, "%Y")))
```

```{r select_hoc_sample}
hoc <- hoc |>
  select(date, year, speaker, party, agenda, terms, text) 
```

## Create a corpus, tokens, and DFM

We first create a `quanteda` corpus. Then we tokenize the speeches and create a document-feature matrix.

```{r create_hoc_corpus}
hoc_corpus <- corpus(
  hoc,
  text_field = "text"
)

ndoc(hoc_corpus)
```

For the sentiment dictionary, we also compound simple negation patterns such as `not good`. This is a simplified approach, but it demonstrates that dictionary methods often require preprocessing decisions.

```{r tokenize_hoc_corpus}
hoc_tokens <- tokens(
  hoc_corpus,
  what = "word",
  remove_punct = TRUE,
  remove_symbols = TRUE,
  remove_numbers = FALSE,
  remove_url = TRUE,
  remove_separators = TRUE,
  split_hyphens = FALSE,
  padding = FALSE
) |>
  tokens_tolower()

hoc_tokens <- tokens_compound(
  hoc_tokens,
  pattern = phrase("not *")
)

hoc_dfm <- dfm(hoc_tokens)

dim(hoc_dfm)

topfeatures(
  dfm_select(hoc_dfm, pattern = "not_*", valuetype = "glob"),
  n = 10
)
```

## Counts and rates

House of Commons speeches differ greatly in length. If we only count dictionary hits, longer speeches have more opportunities to match dictionary terms. For most comparisons, we therefore use a rate:

`100 * dictionary hits / number of tokens`

This gives us dictionary hits per 100 tokens.

```{r inspect_document_lengths}
docvars(hoc_dfm) |>
  summarise(
    min_terms = min(terms),
    median_terms = median(terms),
    max_terms = max(terms)
  )
```

## Off-the-shelf dictionaries

The `quanteda.sentiment` package contains several off-the-shelf sentiment dictionaries. We will use:

- the Lexicoder Sentiment Dictionary, `data_dictionary_LSD2015`;
- the NRC Word-Emotion Association Lexicon, `data_dictionary_NRC`.

```{r inspect_lsd_dictionary}
summary(data_dictionary_LSD2015)
print(data_dictionary_LSD2015, max_nval = 5)
```

### Lexicoder Sentiment Dictionary

We apply the dictionary using `dfm_lookup()`.

```{r apply_lsd_dictionary}
hoc_lsd <- dfm_lookup(
  hoc_dfm,
  dictionary = data_dictionary_LSD2015
)

head(hoc_lsd)
featnames(hoc_lsd)

docvars(hoc_dfm) |>
  mutate(
    docname = docnames(hoc_dfm),
    neg_positive = as.numeric(hoc_lsd[, "neg_positive"]),
    neg_negative = as.numeric(hoc_lsd[, "neg_negative"])
  ) |>
  filter(neg_positive > 0 | neg_negative > 0) |>
  select(docname, date, speaker, party, agenda, neg_positive, neg_negative) |>
  head(10)

lsd_negation_patterns <- data_dictionary_LSD2015[c(
  "neg_positive",
  "neg_negative"
)] |>
  as.list() |>
  unlist() |>
  gsub(pattern = " ", replacement = "_", fixed = TRUE)

kwic(
  hoc_tokens,
  pattern = lsd_negation_patterns,
  valuetype = "glob",
  window = 4
) |>
  as.data.frame() |>
  select(docname, pre, keyword, post) |>
  head(10)
```

The LSD dictionary includes positive and negative words, as well as negated positive and negated negative words. We create a sentiment score as:

`100 * (positive words - negative words) / total tokens`

```{r create_lsd_sentiment}
docvars(hoc_dfm, "lsd_negative") <- as.numeric(hoc_lsd[, "negative"]) +
  as.numeric(hoc_lsd[, "neg_positive"])

docvars(hoc_dfm, "lsd_positive") <- as.numeric(hoc_lsd[, "positive"]) +
  as.numeric(hoc_lsd[, "neg_negative"])

docvars(hoc_dfm, "lsd_sentiment") <- 100 *
  (docvars(hoc_dfm, "lsd_positive") - docvars(hoc_dfm, "lsd_negative")) /
  ntoken(hoc_dfm)

docvars(hoc_dfm) |>
  select(date, speaker, party, lsd_positive, lsd_negative, lsd_sentiment) |>
  head(10)
```

### NRC sentiment and emotion dictionary

The NRC dictionary contains both positive/negative categories and emotion categories such as fear, trust, anger, and joy.

```{r apply_nrc_dictionary}
hoc_nrc <- dfm_lookup(
  hoc_dfm,
  dictionary = data_dictionary_NRC
)

head(hoc_nrc)
featnames(hoc_nrc)
```

```{r create_nrc_sentiment}
docvars(hoc_dfm, "nrc_negative") <- as.numeric(hoc_nrc[, "negative"])
docvars(hoc_dfm, "nrc_positive") <- as.numeric(hoc_nrc[, "positive"])
docvars(hoc_dfm, "nrc_fear") <- as.numeric(hoc_nrc[, "fear"])

docvars(hoc_dfm, "nrc_sentiment") <- 100 *
  (docvars(hoc_dfm, "nrc_positive") - docvars(hoc_dfm, "nrc_negative")) /
  ntoken(hoc_dfm)

docvars(hoc_dfm, "nrc_fear_rate") <- 100 *
  docvars(hoc_dfm, "nrc_fear") /
  ntoken(hoc_dfm)

docvars(hoc_dfm) |>
  select(date, speaker, party, nrc_sentiment, nrc_fear_rate) |>
  head(10)
```

## Comparing sentiment dictionaries

The two dictionaries are intended to measure related constructs, but they are not identical. Let us inspect the correlation and scatterplot.

```{r compare_sentiment_dictionaries}
cor(
  docvars(hoc_dfm, "lsd_sentiment"),
  docvars(hoc_dfm, "nrc_sentiment"),
  use = "complete.obs"
)

sentiment_comparison <- ggplot(
  docvars(hoc_dfm),
  aes(x = lsd_sentiment, y = nrc_sentiment)
) +
  geom_point(alpha = 0.35) +
  labs(
    x = "LSD sentiment",
    y = "NRC sentiment",
    title = "Comparing two off-the-shelf sentiment dictionaries"
  ) +
  theme_minimal()

sentiment_comparison
```

The correlation may be positive, but disagreement is substantively important. It tells us that dictionary choice matters.

## Sentiment over time

Because the sample is approximately balanced by year, we can use it to create simple yearly summaries. These are teaching examples rather than population estimates.

```{r sentiment_by_year}
sentiment_year <- docvars(hoc_dfm) |>
  group_by(year) |>
  summarise(
    lsd_sentiment = mean(lsd_sentiment, na.rm = TRUE),
    nrc_sentiment = mean(nrc_sentiment, na.rm = TRUE),
    nrc_fear_rate = mean(nrc_fear_rate, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(sentiment_year, aes(x = year)) +
  geom_line(aes(y = lsd_sentiment, colour = "LSD")) +
  geom_line(aes(y = nrc_sentiment, colour = "NRC")) +
  labs(
    x = NULL,
    y = "Average sentiment score",
    colour = NULL,
    title = "Dictionary sentiment over time in the HoC sample"
  ) +
  theme_minimal()
```

## A sentence-level example

Dictionary methods can be applied at different units of analysis. Let us select the longest speech in the sample, reshape it to sentences, and inspect the sentence-level NRC fear rate.

```{r longest_speech_sentence_corpus}
longest_speech <- corpus_subset(
  hoc_corpus,
  terms == max(terms)
)

docvars(longest_speech) |>
  select(date, speaker, party, agenda, terms)

longest_speech_sentences <- corpus_reshape(
  longest_speech,
  to = "sentences"
)

ndoc(longest_speech_sentences)
```

```{r sentence_level_fear}
sentence_tokens <- tokens(
  longest_speech_sentences,
  what = "word",
  remove_punct = TRUE,
  remove_symbols = TRUE,
  remove_numbers = FALSE,
  remove_url = TRUE,
  remove_separators = TRUE,
  split_hyphens = FALSE
) |>
  tokens_tolower()

sentence_dfm <- dfm(sentence_tokens)
sentence_nrc <- dfm_lookup(sentence_dfm, dictionary = data_dictionary_NRC)

docvars(sentence_dfm, "sentence") <- seq_len(ndoc(sentence_dfm))
docvars(sentence_dfm, "fear_rate") <- 100 *
  as.numeric(sentence_nrc[, "fear"]) /
  ntoken(sentence_dfm)

ggplot(docvars(sentence_dfm), aes(x = sentence, y = fear_rate)) +
  geom_point(alpha = 0.45) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Sentence",
    y = "NRC fear words per 100 tokens",
    title = "Sentence-level fear in the longest sampled speech"
  ) +
  theme_minimal()
```


## Building a first economic insecurity dictionary

We now turn to the concept from the lecture: **economic insecurity**. By this we mean references to material hardship, financial vulnerability, unstable income, unaffordable living costs, debt burdens, precarious employment, and anxiety about meeting basic needs.

We start with a deliberately simple seed dictionary. It is not meant to be final. It is a first measurement instrument that we can inspect, criticize, and revise.

```{r create_insecurity_dictionary_seed}
insecurity_dictionary_seed <- dictionary(list(
  economic_insecurity = c(
    "poverty", "unemploy*", "inflation*", "wage*", "rent*",
    "bills", "debt*", "cost*", "price*"
  )
))

insecurity_seed_lookup <- dfm_lookup(
  hoc_dfm,
  dictionary = insecurity_dictionary_seed
)

docvars(hoc_dfm, "insecurity_hits_seed") <-
  as.numeric(insecurity_seed_lookup[, "economic_insecurity"])

docvars(hoc_dfm, "insecurity_rate_seed") <- 100 *
  docvars(hoc_dfm, "insecurity_hits_seed") /
  ntoken(hoc_dfm)
```

Let us inspect the speeches with the highest rates. The rate is the number of dictionary hits per 100 tokens.

```{r inspect_top_insecurity_seed}
top_insecurity_seed <- docvars(hoc_dfm) |>
  select(
    date, speaker, party, agenda, terms,
    insecurity_hits_seed, insecurity_rate_seed
  ) |>
  arrange(desc(insecurity_rate_seed))

top_insecurity_seed |>
  head(12)
```

## Rates are not validation

High dictionary rates are useful for finding speeches to inspect. They do not prove that the dictionary measures economic insecurity well.

The next step is to read examples. We will inspect a likely true positive and a likely false positive.

```{r inspect_selected_insecurity_speeches}
hoc |>
  filter(
    (date == as.Date("1986-04-17") & agenda == "Unemployment Costs") |
      (date == as.Date("2024-05-16") & speaker == "Penny Mordaunt")
  ) |>
  select(date, speaker, party, agenda, text)
```

In the 1986 example, terms such as `costs` and `unemployment` are clearly relevant to economic insecurity. In the 2024 example, `Bills` refers to proposed legislation rather than household bills. This is a false positive.

## Inspecting context with KWIC

Keyword-in-context output helps us inspect whether a dictionary term is behaving as intended. Let us look at `bills`.

```{r kwic_bills}
kwic(
  hoc_tokens,
  pattern = "bills",
  window = 10
) |>
  head(30)
```

Some uses of `bills` refer to household costs, but others refer to parliamentary bills. That means `bills` is too ambiguous as a single-word dictionary entry.

## Adapting the dictionary

One way to adapt the dictionary is to remove the ambiguous single-word term `bills` and add more specific multiword phrases such as `energy bills`, `household bills`, and `council tax bills`.

To count multiword expressions in `quanteda`, we first compound them in the token object.

```{r compound_insecurity_phrases}
insecurity_phrases <- phrase(c(
  "cost of living",
  "energy bills",
  "household bills",
  "council tax bills",
  "food banks",
  "zero hours",
  "universal credit"
))

hoc_tokens_insecurity <- tokens_compound(
  hoc_tokens,
  pattern = insecurity_phrases
)

hoc_dfm_insecurity <- dfm(hoc_tokens_insecurity)
```

```{r create_insecurity_dictionary_adapted}
insecurity_dictionary_adapted <- dictionary(list(
  economic_insecurity = c(
    "poverty", "unemploy*", "inflation*", "wage*", "rent*",
    "debt*", "cost*", "price*", "afford*", "arrear*",
    "cost_of_living", "energy_bills", "household_bills",
    "council_tax_bills", "food_banks", "zero_hours",
    "universal_credit"
  )
))

insecurity_adapted_lookup <- dfm_lookup(
  hoc_dfm_insecurity,
  dictionary = insecurity_dictionary_adapted
)

docvars(hoc_dfm_insecurity, "insecurity_hits_adapted") <-
  as.numeric(insecurity_adapted_lookup[, "economic_insecurity"])

docvars(hoc_dfm_insecurity, "insecurity_rate_adapted") <- 100 *
  docvars(hoc_dfm_insecurity, "insecurity_hits_adapted") /
  ntoken(hoc_dfm_insecurity)
```

## Comparing the first and adapted dictionaries

```{r compare_seed_and_adapted_dictionary}
dictionary_comparison <- tibble(
  date = docvars(hoc_dfm, "date"),
  speaker = docvars(hoc_dfm, "speaker"),
  party = docvars(hoc_dfm, "party"),
  agenda = docvars(hoc_dfm, "agenda"),
  seed_rate = docvars(hoc_dfm, "insecurity_rate_seed"),
  adapted_rate = docvars(hoc_dfm_insecurity, "insecurity_rate_adapted")
) |>
  mutate(change = adapted_rate - seed_rate)

dictionary_comparison |>
  arrange(desc(abs(change))) |>
  head(12)
```

The `change` column shows how much the adapted dictionary changes the rate. Positive values mean that the adapted dictionary captures additional insecurity language, for example phrases such as `cost_of_living`, `food_banks`, or `universal_credit`. Negative values mean that the adapted dictionary is more conservative, often because broad or ambiguous terms such as `bills` are no longer counted automatically. The adapted dictionary is not automatically better, but it makes the measurement choices more explicit.

## Economic insecurity over time

Because the sample is approximately balanced by year, we can inspect the average adapted dictionary rate over time. Treat this as a teaching example, not a population estimate.

```{r plot_insecurity_over_time}
insecurity_year <- docvars(hoc_dfm_insecurity) |>
  group_by(year) |>
  summarise(
    insecurity_rate = mean(insecurity_rate_adapted, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(insecurity_year, aes(x = year, y = insecurity_rate)) +
  geom_line() +
  geom_smooth(se = FALSE, span = 0.25) +
  labs(
    x = NULL,
    y = "Dictionary hits per 100 tokens",
    title = "Economic insecurity dictionary over time"
  ) +
  theme_minimal()
```

## Static embeddings for candidate terms

In the lecture, we discussed static embeddings as one way to expand a dictionary. The idea is simple: start with seed terms, find words that appear in similar contexts, and then inspect them as possible additions.

We will not train a new embedding model here, because we did that in Lab 3. The important point for dictionary construction is the workflow:

1. Start with seeds such as `poverty`, `unemployment`, `inflation`, and `debt`.
2. Use embeddings to find nearby terms.
3. Treat neighbours as candidate terms.
4. Inspect KWIC examples before adding anything to the dictionary.

**Exercise:** choose two seed terms from the economic insecurity dictionary. Based on your substantive knowledge, write down three additional terms that might appear in similar contexts. Then inspect those terms with `kwic()`.

```{r exercise_embedding_candidates}
# Example candidate terms to inspect:
candidate_terms <- c("hardship", "afford", "arrears")

kwic(
  hoc_tokens,
  pattern = candidate_terms,
  window = 8
) |>
  head(30)
```

## LLM-assisted dictionary development

An LLM can also help generate candidate dictionary terms. A strong prompt starts from a seed list and asks the model to structure its suggestions.

```text
You are helping build a dictionary for economic insecurity in UK
parliamentary speeches: hardship, unstable income, unaffordable living
costs, debt, precarious work, and anxiety about basic needs.

Start from: poverty, unemployment, inflation, wages, rent, bills, debt.

Suggest additional terms for: income/wages, employment insecurity,
cost of living, housing insecurity, debt, poverty/deprivation,
welfare insecurity, and general hardship language.

Return a table: category | terms/phrases | why it fits | ambiguity or
validation notes. Include UK policy terms, multiword phrases, wildcards,
and likely false positives.
```

The output from this prompt should be treated as a hypothesis. You still need to inspect examples in the corpus and validate the dictionary against human judgment.

## A small validation exercise

Suppose we hand-code 20 speeches after applying the adapted economic insecurity dictionary:

|  | Human relevant | Human not relevant |
|---|---:|---:|
| Dictionary hits | 5 | 3 |
| Dictionary misses | 4 | 8 |

**Exercise:** calculate precision and recall.

```{r exercise_precision_recall}
# Your answer here
```

## Practice exercises

1. Inspect the top 20 speeches according to `insecurity_rate_adapted`. Which look like true positives? Which look doubtful?

```{r exercise_top_insecurity_validation}
# Your answer here
```

2. Use `kwic()` to inspect three ambiguous terms from the adapted dictionary, such as `cost`, `price`, and `rent`. For each term, note whether it usually captures economic insecurity.

```{r exercise_kwic_ambiguous_terms}
# Your answer here
```

3. Add three terms or phrases to `insecurity_dictionary_adapted`, apply the revised dictionary, and compare the top speeches before and after the change.

```{r exercise_revise_insecurity_dictionary}
# Your answer here
```

4. Write a short validation plan for the economic insecurity dictionary. Include how you would sample speeches, what counts as a true positive, and how you would report changes to the dictionary.

```{r exercise_reflection}
# Your notes here
```
