---
title: "QTA Lab 01: Introduction to R and the House of Commons Corpus"
format:
  gfm: default
editor: source
execute:
  echo: true
  message: false
  warning: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  results = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

## Learning goals

In this first lab, you will learn how to:

- use Positron to run R code;
- create and inspect basic R objects;
- distinguish numeric, character, logical, and date variables;
- load R packages;
- read in and inspect a small sample of House of Commons speeches;
- treat text as data using string operations;
- make a first plot from political text data.

Throughout the course, we will use a sample of approximately 5,000 House of Commons speeches from 1945 to 2025. Each row in the dataset is a speech. The original dataset is much larger (and ranges from 1919-2025) and was put together by Stefan Couperus and Martijn Schoonvelde, but we will use a smaller sample for teaching purposes.

## Positron

We will use `R` in Positron. Positron is an integrated development environment, or IDE, for data science work. It helps you write code, run code, inspect objects, view plots, and work with files in one place. Positron also allows you to switch between R and Python, and to use R Markdown and Quarto documents for reproducible research. If you have used other IDEs, such as RStudio, you will find Positron familiar, but otherwise it may take some getting used to (but you will!). Positron is free to use and available for Windows, Mac, and Linux. You can download it from [here](https://positron.posit.co/download.html). 

For this course, open the main course folder in Positron rather than opening individual lab folders. This folder is a Quarto project, which means that code is evaluated relative to the main course folder when the labs are rendered. Using one project folder also keeps file paths simple: paths such as `Labs/Lab_1/hc_sample_1945_2025.rds` start from the main course folder.

The interface will usually contain these areas:

- Activity bar, where you can switch between the file explorer, search, source control, and extensions.
- Primary side bar, where you can navigate files and folders, search for text, and manage source control.
- The Console, where you can run R commands directly.
- The Editor, where you write and save code in files such as `.R` and `.qmd` files.
- The Variables or Environment pane, where you can inspect objects you have created.
- The Plots, Files, Help, or Viewer panes, where you can inspect output, navigate files, and read documentation.

I recommend you take some time to familiarise yourself with Positron. You can find more information about Positron in the [Positron documentation](https://positron.posit.co/docs/).

## Running code

Each lab file is a Quarto document, which is a type of R Markdown file. You can run code in a Quarto document in several ways. You can run the whole document interactively by clicking the **Run all** button at the top of the editor. This runs the code chunks in your current R session. You can also run individual code chunks by clicking the **Run** button attached to each chunk. To create the final rendered lab document, use **Render**. You can run one line of code from a `.qmd` file by placing your cursor on that line and pressing `Cmd + Enter` on Mac or `Ctrl + Enter` on Windows.

In these lab documents, code appears in grey blocks called code chunks. Lines that begin with `#` are comments. They are not executed by R, but they explain what the code is doing.

## R as a calculator

Let's start very simply: R can be used as a calculator. This is a useful way to get comfortable with the Console.

```{r calculator}
# Addition
2 + 2

# Exponentiation
3^2

# Subtraction
4 - 5

# Roughly how many speeches per year are in our teaching sample?
5000 / 81
```

R also contains many built-in functions. Functions take inputs, called arguments, and return outputs. For example, the `mean()` function takes a numeric vector as input and returns the mean of that vector. The `median()` function returns the median, and the `sqrt()` function returns the square root. In the code below, `c()` is a function that creates a vector from the numbers inside the parentheses.

```{r built_in_functions}
mean(c(2, 4))
median(c(2, 4, 100))
sqrt(81)
```

## Objects

We can save results as objects using the assignment operator `<-`.

As you will see later, the House of Commons sample contains a variable called `terms`, which gives the number of terms in each speech. Let us make a small vector of speech lengths by hand from five speeches.

```{r create_numeric_vector}
speech_terms <- c(72, 53, 29, 30, 50)
speech_terms
```

Once the object exists, we can use it in later commands.

```{r inspect_numeric_vector}
mean(speech_terms)
max(speech_terms)
min(speech_terms)
```

R objects can contain different types of data. Text is usually stored as a character vector. Let's say we have the names of five speakers in the sample. We can create a character vector using `c()`.

```{r create_character_vector}
speakers <- c(
  "Mr. Johnstone",
  "Mr. Woodburn",
  "Mr. Eden",
  "Vice-Admiral Taylor",
  "Sir M. Sueter"
)

speakers
class(speakers)
```

What happens if we try to calculate the mean of a vector of speaker names?

```{r mean_character_vector}
# mean(speakers)
```

The line is commented out because it will produce an error. That error is useful: it tells us that `mean()` expects numeric input, but speaker names are character data.

## Data types

The most common data types we will use in this course are:

- `numeric`: numbers, such as `72` or `196.6`;
- `integer`: whole numbers, such as a speech number;
- `character`: text, such as a speaker name or speech;
- `logical`: `TRUE` or `FALSE`;
- `Date`: calendar dates.

You can inspect the type of an object using `class()`.

```{r inspect_data_types}
speech_date <- as.Date("1945-01-16")
mentions_trade <- TRUE

class(speech_terms)
class(speakers)
class(speech_date)
class(mentions_trade)
```

## Packages

Base R contains many functions, but much of the R ecosystem is organised around packages. Packages are collections of functions, data, and documentation that have been produced by the R community. You can install packages from CRAN, the Comprehensive R Archive Network, or from other sources such as GitHub. Once a package is installed, you can load it into your R session using `library()`.

Today we will use three packages:

- `dplyr` for working with data frames;
- `stringr` for working with text strings;
- `ggplot2` for plotting.

You only need to install a package once, but you need to load it every time you start a new R session.

```{r install_packages}
# install.packages("dplyr")
# install.packages("stringr")
# install.packages("ggplot2")
```

```{r load_packages}
library(dplyr)
library(stringr)
library(ggplot2)
```

It is good practice to cite packages used in an analysis. You can see how to cite a package using `citation()`.

```{r cite_package}
citation("stringr")
```

## Read the House of Commons sample

The full House of Commons dataset is large (6,821,869 speeches). For teaching purposes, we will use a smaller sample covering 1945 to 2025.

This sample has already been cleaned in two ways. First, very short speeches and chair contributions have been removed. Second, common party label variants have been standardised. The original party label is still available in `party_raw`.

The data file is stored as an `.rds` file, which is a native R file format. Here we use `readRDS()` to read the file. If our data would have been stored as a CSV file, we would have used `read.csv()` (base R) or `read_csv()` (tidyverse) instead.

Because this course folder is a Quarto project, both rendering and interactive work should use the main course folder as the working directory. The path below therefore starts from the main course folder.

```{r read_hoc_sample}
data_path <- "Labs/Lab_1/hc_sample_1945_2025.rds"

hoc <- readRDS(data_path)
```

Let us inspect the object. The `str()` function is useful when you first read in a dataset because it shows the variables, their data types, and a few example values.

```{r inspect_hoc_sample}
class(hoc)
dim(hoc)
names(hoc)
str(hoc)
```

The object is a data frame. A data frame is a table where rows are observations and columns are variables.

The main variables are:

- `date`: date of the speech;
- `agenda`: agenda item or debate heading;
- `speaker`: speaker name;
- `party`: standardised party label;
- `party_raw`: original party label;
- `terms`: number of terms in the speech;
- `text`: speech text;
- `speech_url`: link to the original parliamentary record.

In the next code chunk, we select a few variables and show the first five rows of the data frame. The "|>" operator is called the pipe. It takes the output of one command and passes it as input to the next command. So the code below can be read as follows: "Take the `hoc` data frame, select the `date`, `speaker`, `party`, `agenda`, `terms`, and `text` variables, and then show the first five rows."

```{r preview_hoc_sample}
hoc |>
  select(date, speaker, party, agenda, terms, text) |>
  head(5)
```

## Inspecting the corpus

We can start with simple descriptive questions. How many speeches are in the sample? What dates does it cover? Which parties appear most often?

```{r corpus_overview}
nrow(hoc)
range(hoc$date)
sort(table(hoc$party), decreasing = TRUE)
```

Because this is a sample, these counts should not be interpreted as the true number of speeches made by each party in Parliament. They are useful for learning how to work with text data.

Let us add a year variable and a character-based speech length variable. The code below uses `mutate()` to create new variables. The `format()` function extracts the year from the `date` variable, and `nchar()` counts the number of characters in the speech text.

```{r create_hoc_variables}
hoc <- hoc |>
  mutate(
    year = as.integer(format(date, "%Y")),
    speech_length_chars = nchar(text)
  )

hoc |>
  select(date, year, speaker, party, terms, speech_length_chars) |>
  head(10)
```

## Text as character data

In R, text is usually stored as character data. The `text` column in our data frame contains the speech itself.

```{r inspect_speech_text}
speech_text <- hoc$text[1:5]

speech_text
class(speech_text)
```

The `stringr` package gives us useful functions for working with text.

```{r string_operations}
# Count characters in each speech
str_length(speech_text)

# Convert text to lower case
str_to_lower(speech_text[1])

# Show the first 120 characters of each speech
str_sub(speech_text, start = 1, end = 120)
```

We can also detect whether a pattern appears in a speech. This returns a logical vector: `TRUE` when the pattern is found and `FALSE` when it is not found.

```{r detect_words}
str_detect(str_to_lower(speech_text), "trade")
str_detect(str_to_lower(speech_text), "war")
```

In political text analysis, this kind of pattern detection is a building block for more elaborate methods. The difficult part is usually not counting words. The difficult part is deciding what the words mean in context.

## A first text-based variable

Suppose we want to know which speeches mention the economy. We can create a logical variable for this. The code below uses `mutate()` to create three new variables: `mentions_economy`, `mentions_nhs`, and `mentions_europe`. Each variable is `TRUE` if the speech contains one of the relevant keywords, and `FALSE` otherwise. `str_detect()` is used to detect the keywords, and `str_to_lower()` converts the speech text to lower case so that the search is case-insensitive. The `\\b` in the regular expression indicates a word boundary, so that we only match whole words. We'll learn more about regular expressions later in the course so don't worry if you don't understand them yet.

```{r create_topic_indicators}
hoc <- hoc |>
  mutate(
    mentions_economy = str_detect(
      str_to_lower(text),
      "\\beconomy\\b|\\beconomic\\b|\\beconomics\\b"
    ),
    mentions_nhs = str_detect(
      str_to_lower(text),
      "\\bnhs\\b|national health service"
    ),
    mentions_europe = str_detect(
      str_to_lower(text),
      "\\beurope\\b|\\beuropean\\b"
    )
  )

hoc |>
  summarise(
    economy_speeches = sum(mentions_economy),
    nhs_speeches = sum(mentions_nhs),
    europe_speeches = sum(mentions_europe)
  )
```

This is a very simple measurement strategy. It will miss some relevant speeches and include some irrelevant ones. Later in the course, we will discuss dictionaries, supervised classification, embeddings, and LLM-assisted annotation as more flexible ways to measure concepts in text.

## Plotting data

We can use `ggplot2` to plot variables from the data frame. First, let us plot the most common parties in the sample.

The code below counts the number of speeches by party, filters to parties with at least 10 speeches, and then creates a horizontal bar plot. The `reorder()` function is used to order the bars by count.

```{r plot_speeches_by_party}
party_counts <- hoc |>
  count(party, sort = TRUE) |>
  filter(n >= 10)

party_plot <- ggplot(
  party_counts,
  aes(x = reorder(party, n), y = n)
) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Number of speeches",
    title = "Party labels in the House of Commons teaching sample"
  ) +
  theme_minimal()

party_plot
```

We can also inspect how mentions of selected topics vary over time in the sample. The `group_by()` and `summarise()` functions are used to calculate the average share of speeches mentioning each topic by year. We then use `.groups = "drop"` to remove the grouping structure from the resulting data frame. Finally, we use `ggplot2` to create a line plot showing the share of speeches mentioning each topic over time.

```{r plot_topic_mentions}
topic_by_year <- hoc |>
  group_by(year) |>
  summarise(
    economy = mean(mentions_economy),
    nhs = mean(mentions_nhs),
    europe = mean(mentions_europe),
    .groups = "drop"
  )

topic_plot <- ggplot(topic_by_year, aes(x = year)) +
  geom_line(aes(y = economy, colour = "Economy")) +
  geom_line(aes(y = nhs, colour = "NHS")) +
  geom_line(aes(y = europe, colour = "Europe")) +
  scale_y_continuous(labels = function(x) paste0(round(100 * x), "%")) +
  labs(
    x = NULL,
    y = "Share of sampled speeches",
    colour = NULL,
    title = "Simple keyword mentions over time"
  ) +
  theme_minimal()

topic_plot
```

Finally, let us save one of the plots. Because our paths start from the main course folder, we write the PNG file to the `Labs/Lab_1` folder explicitly.

```{r save_party_plot}
ggsave("Labs/Lab_1/hoc_party_counts.png", plot = party_plot, width = 8, height = 5)
```

## A note on LLMs and text input

Large language models can classify, summarise, and extract information from text. But they still need carefully designed inputs. Before using an LLM, we need to decide what information to provide.

Consider one speech from the sample:

```{r inspect_one_speech}
one_speech <- hoc |>
  select(date, speaker, party, agenda, text) |>
  slice(1)

one_speech
```

If we asked an LLM to classify the topic of this speech, should the prompt include the `speaker`, `party`, `date`, and `agenda`, or only the speech text? There is no universal answer. Metadata may help the model interpret the speech, but it may also introduce bias into the classification.

We will return to these questions later in the course. For now, the important point is that LLM-based text analysis still depends on the same research design choices as other QTA methods: unit of analysis, corpus construction, measurement, validation, and documentation.

## Practice exercises

1. Create a vector called `parties` that contains the first ten party labels in the HoC sample. Inspect its class and length.

```{r exercise_parties_vector}
# Your answer here
```

2. Create a new variable called `mentions_immigration` that detects speeches containing either `immigration`, `immigrant(s)`, or `migrant(s)`.

```{r exercise_immigration_indicator}
# Your answer here
```

3. Count how many speeches mention immigration by party. Display only parties with at least 10 speeches in the sample.

```{r exercise_immigration_by_party}
# Your answer here
```

4. Make a plot showing the average number of terms per speech for the six most common parties in the sample.

```{r exercise_average_terms_plot}
# Your answer here
```

5. Pick one speech from the dataset. Write down what metadata you would include if you wanted to ask an LLM to classify the speech topic, and explain why.

```{r exercise_llm_metadata}
# Your notes here
```
